Shapes analysis for time series.

### 2024-05-22 NEURIPS24/appendix.tex
update

First, we show the model identifiability when the kernel $K_G$ is well specified: the estimated parameter is a good approximation of the generating parameter when the generation and the estimation procedure use the same hyperparameters for the RKHS kernel $K_\msg$.
All the hyperparameter values for generation and estimation are given in \Cref{appendix:numerics_synthetic}.
We fix the initial control points as $\msx=\left(x_k=(k,\sin(2\pi k/300))\right)_{k\in[300]} $.
Given $m_{s}\in \Nset_{>0}$ and $t_{a},s_{a}>0$, we randomly generate initial momentums $\mathbf{\alpha}^*=(\alpha_k^*)_{k\in[\mathbf{n}_0]}$ with the following sampling, called Gen($m_s,t_a,s_a$):
For any $k\in[\mathbf{n}_0]$, $\mathbf{\alpha}_k'$ is sampled according to a Gaussian normal distribution $\mathcal{N}(0_{d+1},I_{d+1})$.
Then, $(\alpha_k')_{k\in[\mathbf{n}_0]}$ is regularized by a rolling average of size $m_{s}$, we get $\bar{\mathbf{\alpha}}'=(\bar{\alpha}_k')_{k\in[\mathbf{n}_0]}$.
Finally, we normalize $\bar{\mathbf{\alpha}}'$ to derive $\mathbf{\alpha}^*$ such that $|([\alpha_k^*]_t)_{k\in[\mathbf{n}_0]}|=t_{\text{amp}}$ and $|([\alpha_k^*]_s)_{k\in[\mathbf{n}_0]}|=s_{\text{amp}}$ for any $k\in[\mathbf{n}_0]$, denoting by $[\alpha_k^*]_t,[\alpha_k^*]_s$ the time and space coordinates of $\alpha_k^*$ respectively.
Note that the regularizing step $(\mathbf{\alpha}_k')_{k\in[\mathbf{n}_0]}\to \bar{\mathbf{\alpha}}' $ is necessary to obtain realistic deformations which take into account the regularity induced by the RKHS $\msv$.

@@

\subsection{Appendix for experiment: Classification benchmark on regularly sampled datasets}

@@

\paragraph{Methods} We compare TS-LDDMM with a method from function~\cite{wu2024shape}

\paragraph{Protocole}

>>>>

First, we show the model identifiability when the kernel $K_G$ is well specified: the estimated parameter is a good approximation of the generating parameter when the generation and the estimation procedure use the same hyperparameters for the RKHS kernel $K_\msg$.
`%All the hyperparameter values for generation and estimation are given in \Cref{appendix:numerics_synthetic}.
We fix the initial control points as $\msx=\left(x_k=(k,\sin(2\pi k/300))\right)_{k\in[300]} $.
Given $m_{s}\in \Nset_{>0}$ and $t_{a},s_{a}>0$, we randomly generate initial momentums $\mathbf{\alpha}^*=(\alpha_k^*)_{k\in[\mathbf{n}_0]}$ with the following sampling, called Gen($m_s,t_a,s_a$):
For any $k\in[\mathbf{n}_0]$, $\mathbf{\alpha}_k'$ is sampled according to a Gaussian normal distribution $\mathcal{N}(0_{d+1},I_{d+1})$.
Then, $(\alpha_k')_{k\in[\mathbf{n}_0]}$ is regularized by a rolling average of size $m_{s}$, we get $\bar{\mathbf{\alpha}}'=(\bar{\alpha}_k')_{k\in[\mathbf{n}_0]}$.
Finally, we normalize $\bar{\mathbf{\alpha}}'$ to derive $\mathbf{\alpha}^*$ such that $|([\alpha_k^*]_t)_{k\in[\mathbf{n}_0]}|=t_{\text{amp}}$ and $|([\alpha_k^*]_s)_{k\in[\mathbf{n}_0]}|=s_{\text{amp}}$ for any $k\in[\mathbf{n}_0]$, denoting by $[\alpha_k^*]_t,[\alpha_k^*]_s$ the time and space coordinates of $\alpha_k^*$ respectively.
Note that the regularizing step $(\mathbf{\alpha}_k')_{k\in[\mathbf{n}_0]}\to \bar{\mathbf{\alpha}}' $ is necessary to obtain realistic deformations which take into account the regularity induced by the RKHS $\msv$.

@@

\section{Appendix for experiment: Classification benchmark on regularly sampled datasets}

@@

%\paragraph{Methods} We compare TS-LDDMM with a method from function~\cite{wu2024shape}

%\paragraph{Protocole}

### 2024-05-22 NEURIPS24/checklist.tex
update

    \item {\bf Experimental Result Reproducibility}
    \item[] Question: Does the paper fully disclose all the information needed to reproduce the main experimental results of the paper to the extent that it affects the main claims and/or conclusions of the paper (regardless of whether the code and data are provided or not)?
    \item[] Answer: \answerYes{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
    \item[] Justification: The optimization methodology is described in \Cref{section:methodology} and all numerical details and protocol are given in \Cref{section:experiments} and in \Cref{appendix:classification_implementation}-\ref{appendix:optimizers_details}-\ref{appendix:numerics_synthetic}-\ref{appendix:identifiability}.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer NA means that the paper does not include experiments.
        \item If the paper includes experiments, a No answer to this question will not be perceived well by the reviewers: Making the paper reproducible is important, regardless of whether the code and data are provided or not.
        \item If the contribution is a dataset and/or model, the authors should describe the steps taken to make their results reproducible or verifiable. 
        \item Depending on the contribution, reproducibility can be accomplished in various ways. For example, if the contribution is a novel architecture, describing the architecture fully might suffice, or if the contribution is a specific model and empirical evaluation, it may be necessary to either make it possible for others to replicate the model with the same dataset, or provide access to the model. In general. releasing code and data is often one good way to accomplish this, but reproducibility can also be provided via detailed instructions for how to replicate the results, access to a hosted model (e.g., in the case of a large language model), releasing of a model checkpoint, or other means that are appropriate to the research performed.
        \item While NeurIPS does not require releasing code, the conference does require all submissions to provide some reasonable avenue for reproducibility, which may depend on the nature of the contribution. For example
        \begin{enumerate}
            \item If the contribution is primarily a new algorithm, the paper should make it clear how to reproduce that algorithm.
            \item If the contribution is primarily a new model architecture, the paper should describe the architecture clearly and fully.
            \item If the contribution is a new model (e.g., a large language model), then there should either be a way to access this model for reproducing the results or a way to reproduce the model (e.g., with an open-source dataset or instructions for how to construct the dataset).
            \item We recognize that reproducibility may be tricky in some cases, in which case authors are welcome to describe the particular way they provide for reproducibility. In the case of closed-source models, it may be that access to the model is limited in some way (e.g., to registered users), but it should be possible for other researchers to have some path to reproducing or verifying the results.
        \end{enumerate}
    \end{itemize}

@@

\item {\bf Experimental Setting/Details}
    \item[] Question: Does the paper specify all the training and test details (e.g., data splits, hyperparameters, how they were chosen, type of optimizer, etc.) necessary to understand the results?
    \item[] Answer: \answerYes{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
    \item[] Justification: The experiments setting is clearly presented in \Cref{section:experiments} for the experiments on the mouse dataset. The full details of all experiments are presented in \Cref{appendix:classification_implementation}-\ref{appendix:optimizers_details}-\ref{appendix:numerics_synthetic}-\ref{appendix:identifiability}.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer NA means that the paper does not include experiments.
        \item The experimental setting should be presented in the core of the paper to a level of detail that is necessary to appreciate the results and make sense of them.
        \item The full details can be provided either with the code, in appendix, or as supplemental material.
    \end{itemize}

>>>>

    \item {\bf Experimental Result Reproducibility}
    \item[] Question: Does the paper fully disclose all the information needed to reproduce the main experimental results of the paper to the extent that it affects the main claims and/or conclusions of the paper (regardless of whether the code and data are provided or not)?
    \item[] Answer: \answerYes{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
    \item[] Justification: The optimization methodology is described in \Cref{section:methodology} and all numerical details and protocol are given in \Cref{section:experiments} and in \Cref{appendix: settings_classification}-\ref{appendix:optimizers_details}-\ref{appendix:numerics_synthetic}-\ref{appendix: settings_classification}.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer NA means that the paper does not include experiments.
        \item If the paper includes experiments, a No answer to this question will not be perceived well by the reviewers: Making the paper reproducible is important, regardless of whether the code and data are provided or not.
        \item If the contribution is a dataset and/or model, the authors should describe the steps taken to make their results reproducible or verifiable. 
        \item Depending on the contribution, reproducibility can be accomplished in various ways. For example, if the contribution is a novel architecture, describing the architecture fully might suffice, or if the contribution is a specific model and empirical evaluation, it may be necessary to either make it possible for others to replicate the model with the same dataset, or provide access to the model. In general. releasing code and data is often one good way to accomplish this, but reproducibility can also be provided via detailed instructions for how to replicate the results, access to a hosted model (e.g., in the case of a large language model), releasing of a model checkpoint, or other means that are appropriate to the research performed.
        \item While NeurIPS does not require releasing code, the conference does require all submissions to provide some reasonable avenue for reproducibility, which may depend on the nature of the contribution. For example
        \begin{enumerate}
            \item If the contribution is primarily a new algorithm, the paper should make it clear how to reproduce that algorithm.
            \item If the contribution is primarily a new model architecture, the paper should describe the architecture clearly and fully.
            \item If the contribution is a new model (e.g., a large language model), then there should either be a way to access this model for reproducing the results or a way to reproduce the model (e.g., with an open-source dataset or instructions for how to construct the dataset).
            \item We recognize that reproducibility may be tricky in some cases, in which case authors are welcome to describe the particular way they provide for reproducibility. In the case of closed-source models, it may be that access to the model is limited in some way (e.g., to registered users), but it should be possible for other researchers to have some path to reproducing or verifying the results.
        \end{enumerate}
    \end{itemize}

@@

\item {\bf Experimental Setting/Details}
    \item[] Question: Does the paper specify all the training and test details (e.g., data splits, hyperparameters, how they were chosen, type of optimizer, etc.) necessary to understand the results?
    \item[] Answer: \answerYes{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
    \item[] Justification: The experiments setting is clearly presented in \Cref{section:experiments} for the experiments on the mouse dataset. The full details of all experiments are presented in \Cref{appendix: benchmark robustness}-\ref{appendix:optimizers_details}-\ref{appendix:numerics_synthetic}-\ref{appendix: settings_identifiability}.
    \item[] Guidelines:
    \begin{itemize}
        \item The answer NA means that the paper does not include experiments.
        \item The experimental setting should be presented in the core of the paper to a level of detail that is necessary to appreciate the results and make sense of them.
        \item The full details can be provided either with the code, in appendix, or as supplemental material.
    \end{itemize}

### 2024-05-22 NEURIPS24/experiment.tex
update

For consiceness, several experiments are relagated in appendix:

\begin{enumerate}
    \item \textbf{TS-LDDMM representation identifiability, \Cref{appendix:identifiability}:} On synthetic data, we evaluate the ability of our method to retrieve the parameter $v_0^*$ that encodes the deformation $\varphi^{\{v_0^*\}}$ acting on a time series graph $\msg$ by solving the geodesic shooting problem \eqref{eq:relaxation} between $\msg$ and $\varphi^{\{v_0^*\}}.\msg$.
     \textbf{Results} show that TS-LDDMM representations are identifiable or weakly identifiable depending on the velocity field kernel $K_G$ specification.

    \item \textbf{Robustness to irregular sampling} We compare the robustness of TS-LDDMM representation with 9 URL methods handling irregularly sampled multivariate time series on 15 shape-based datasets (7 univariates \& 8 multivariates).
     We assess methods' classification performances under regular sampling (0\% missing rate) and three irregular sampling regimes (30\%, 50\%, and 70\% missing rates), according to the protocol depicted in \cite{kidger2020neural}.
      \textbf{Results} show that our method, TS-LDDMM, outperforms all methods for sampling regimes with missing rates: 0\%, 30\%, and 50\%.
      %{\color{red} J'enlèverai cette phrase ou je développerai en Appendix : The performance decrease of the Kernel-SVC based on TS-LDDMM representation is sensibly due to the misspecification of its regularization parameter.}

    \item \textbf{Classification benchmark on regularly sampled datasets, \Cref{appendix:classification_shape_analysis}:} We compare performances of a kernel support vector machine (SVC) algorithm based on TS-LDDMM representation with 3 state-of-the-art classification methods from shape analysis on 15 shape-based datasets (7 univariates \& 8 multivariates). \textbf{Results} show that the TS-LDDMM-based method outperforms other methods (best performances over 13 datasets), making TS-LDDMM representation relevant for time series shape analysis.
\end{enumerate}

\vspace{-1ex}
\subsection{Interpretability: analysis of respiratory behavior in mice}
\label{section:interpretability}
\vspace{-1ex}
%We consider a dataset composed of mouse respiratory time series before and after a drug injection.
% A complete presentation of this dataset is given in \Cref{appendix:mouse_dataset}.
%  The mouse are divided in two groups depending on their genotypes : \textit{colq} and \textit{wt}.
%  We aim to study the difference of respiratory shapes according to the genotype.
%   That is why, TS-LDDMM \eqref{eq:general_optimization_problem} is applied to derive the features representations $(\mathbf{\alpha}_j)_{j\in[N]}\in (\Rset^2)^N$ related to $N$=\sam{fill} respiratory time series coming from 14 mouse (7 \textit{colq} and \textit{wt}) before drug injection.
%   Then, a Principal Components Analysis (PCA) is performed on $(\mathbf{\alpha}_j)_{j\in[N]}$
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/ts_ddmm_shooting.pdf}
    \caption{TS-LDDMM}
    \label{fig:ts-lddmm shooting}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/lddmm_shooting.pdf}
    \caption{LDDMM}
    \label{fig:lddmm shooting}
  \end{subfigure}
  \caption{Analysis of the two principal components (PC) related to mice' respiratory cycles before exposure for TS-LDDMM \Cref{fig:ts-lddmm shooting}, and LDDMM \Cref{fig:lddmm shooting}.
  In both cases, I) displays PC densities according to mice genotype and II) displays  deformations of the reference graph $S_0$ along each PC.}
  \label{fig:exp1}
\end{figure*}

\begin{wrapfigure}{r}{0.50\textwidth}
  \centering
  \begin{subfigure}[b]{0.16\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
    \caption{colq cycle}
    \label{fig:colq-cycle}
  \end{subfigure}
  \begin{subfigure}[b]{0.16\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
    \caption{PC1 vs PC2}
    \label{fig: pcs-scatter}
  \end{subfigure}
  \begin{subfigure}[b]{0.16\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
    \caption{wt cycle}
    \label{fig:wt-reference}
  \end{subfigure}
  \caption{\Cref{fig:colq-cycle} is a \textbf{colq} respiratory cycle. \Cref*{fig: pcs-scatter} displays individual mouse reference respiratory cycle in the TS-LDDMM PC1-PC2 coordinates. \Cref{fig:wt-reference} is a referent respiratory cycle of a \textbf{wt} mouse learned with TS-LDDMM. }
  \vspace{-1em}
\end{wrapfigure}

@@

This experiment highlights the \textit{interpretability} of TS-LDDMM representation for studying the inter-individual variability in time series biomedical datasets. We consider a multivariate time series dataset that accounts for mice's nasal and thoracic airflow evolution when exposed to an irritant molecule altering respiratory functions \cite{nervo2019respiratory}.
The dataset is divided into two groups, one composed of 7 control mice (\textbf{wt}) and the other of 7 mice (\textbf{colq}) deficient in an enzyme involved in the control of respiration. For each mouse, airflows were recorded for 15 to 20 minutes before exposure to the irritant molecule and then for 35 to 40 minutes. A complete 
description of the dataset is given in the \Cref{appendix:mouse_dataset}.
\vspace{-1ex}
\paragraph{Result summary.} By comparing the shape of individual respiratory cycles, we show that TS-LDDMM features can encode genotype distinctive breathing behaviors and their evolution after exposure to the irritant molecule contrary to LDDMM \cite{glaunes2008large}. 
\vspace{-1ex}
\paragraph{Protocol.} For both experiments and representations (TS-LDDMM and LDDMM), we derive the reference respiratory cycle's graph $S_0$ and the representations $(\alpha_j)_{j\in[N]}$ related to $N$ respiratory cycles extracted according to the procedure \cite{germain2023unsupervised} by solving  \eqref{eq:general_optimization_problem}.
 Then, we perform a kernel PCA on the initial velocity field $\left(v_0(\alpha_j,S_0)\right)_{j\in[N]}\in \msv^{N_1}$ defined in \eqref{eq:def_v0} to breathing behavior.
  The first experiment includes $N_1 = 700$ respiratory cycles collected before exposure. The second experiment includes $N_2 = 1400$ respiratory cycles with 25\% (resp. 75\%) before (resp. after) exposure.
   \Cref{appendix: mice_exp_setting} describes methods settings.
    Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales.

    \vspace{-1ex}
\paragraph{Breathing behaviors before exposure.}
 We focus on the analysis of the two first Principal Components (PC) for TS-LDDMM (\Cref{fig:ts-lddmm shooting}), and LDDMM (\Cref{fig:lddmm shooting}).
  A respiratory cycle is an inspiration (positive flow) followed by an expiration (negative flow); \Cref{fig:wt-reference} displays an example of \textbf{wt} mouse respiratory cycle.
   \Cref{fig:exp1} shows that principal components learned with TS-LDMM lead to deformations that remain respiratory cycles, while deformations learned with LDDMM are challenging to interpret as respiratory cycles.
   The LDDMM velocity field kernel is a Gaussian anisotropic kernel that accounts for time and space scales; however, the entanglement of time and space dimensions in the kernel does not guarantee the graph structure, and it makes the convergence of the method difficult (relative varifold loss error: TS-LDDMM: 0.06, LDDMM: 0.11).
    Regarding TS-LDDMM, \Cref{fig:ts-lddmm shooting} shows its principal components refer to different types of deformations.
     By interpreting \Cref{fig:ts-lddmm shooting}: PC1 accounts for time warping, PC2 expresses the trade-off between inspiration and expiration duration.
      Compared to \textbf{wt} mice, the distribution of \textbf{colq} mice feature along the PC1 axis has a heavy left tail, and the associated deformation (-2 $\sigma_{\text{PC}}$) shows an inspiration with two peaks.
       \Cref{fig:colq-cycle} shows an example of such respiratory cycles, which may be caused by motor impairment due to enzyme deficiency \cite{germain2023unsupervised}.
        Finally, \Cref{fig: pcs-scatter} shows that PC1 and PC2 capture the main differences between the two groups as their respective reference graphs $S^j$ are located in different parts of the space.

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.95\linewidth]{"./pictures/exp2.pdf"}
  \caption{Analysis of the first Principal Component (PC1) related to mice' respiratory cycles before and after exposure for TS-LDDMM. Figure A) displays PC densities according to mice genotype, Figure B) displays  deformations of the reference graph $S_0$ along PC1, and Figure C) displays respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCA}
  \vspace{-1.5em}
\end{figure*}
\vspace{-1ex}
\paragraph{Breathing behaviors' evolution after exposure to the irritant molecule} 
TS-LDDMM representation features are learned on a dataset that includes respiratory cycles before and after exposure.
 \Cref{fig:exp_2_PCA} displays the first principal component PC since it encodes the effect of the irritant molecule as depicted in \Cref{fig:exp_2_PCA}-C) (the exposure occurs at 20 minutes). \Cref{fig:exp_2_PCA}-B) shows that the deformation (-1.5 $\sigma_{\text{PC}}$) leads to longer respiratory cycles that include a pause between inspiration and expiration as observed in \cite{germain2023unsupervised}.
  Conjointly, \Cref{fig:exp_2_PCA}-A) shows a bimodal distribution for \textbf{wt} mice whereas \textbf{wt}' distributions before exposure were unimodal (\Cref{fig:ts-lddmm shooting}). Indeed, the irritant molecule inhibits the action of the deficient enzyme, \textbf{wt} mice strongly react to the irritant molecule, whereas \textbf{colq} mice better adapt due to their deficiency \cite{germain2023unsupervised}.
%Secondly, we compare breathing behaviors before and after exposure to observe the impact of the irritant molecule.
% We follow the same procedure as for before exposure, but we take $N_2=1400$ respiratory cycles extracted according 
% to the procedure CITE....
% In \Cref{fig:exp_2_PCA}, we focus on the first Principal Component (PC) since it encodes the effect of the irritant molecule as demonstrated on Figure \ref{fig:exp_2_PCA}-C) (exposure at time 20).
% We observe on \Cref{fig:exp_2_PCA}-B) that after exposure the mouse have longuer breath such that the expiration is longuer than inspiration.
%  Moreover, we deduce from Figure \ref{fig:exp_2_PCA}-A) that \textbf{colq} are more constant in their breath compared to \textbf{wt} after exposure.
%
% In Figure REF, we 
% also display the reference respiratory cycle S0 and its deformations in the principal component directions.
%  Additionally, we learn each mouse's reference respiratory cycle and represent them in the first and third PC coordinates system in Figure REF.

>>>>

For conciseness, several experiments are relegated in appendix:

\begin{enumerate}
    \item \textbf{TS-LDDMM representation identifiability, \Cref{appendix: settings_identifiability}:} On synthetic data, we evaluate the ability of our method to retrieve the parameter $v_0^*$ that encodes the deformation $\varphi^{\{v_0^*\}}$ acting on a time series graph $\msg$ by solving the geodesic shooting problem \eqref{eq:relaxation} between $\msg$ and $\varphi^{\{v_0^*\}}.\msg$.
     \textbf{Results} show that TS-LDDMM representations are identifiable or weakly identifiable depending on the velocity field kernel $K_G$ specification.

    \item \textbf{Robustness to irregular sampling, \Cref{appendix: benchmark robustness}:} We compare the robustness of TS-LDDMM representation with 9 URL methods handling irregularly sampled multivariate time series on 15 shape-based datasets (7 univariates \& 8 multivariates).
     We assess methods' classification performances under regular sampling (0\% missing rate) and three irregular sampling regimes (30\%, 50\%, and 70\% missing rates), according to the protocol depicted in \cite{kidger2020neural}.
      \textbf{Results} show that our method, TS-LDDMM, outperforms all methods for sampling regimes with missing rates: 0\%, 30\%, and 50\%.
      %{\color{red} J'enlèverai cette phrase ou je développerai en Appendix : The performance decrease of the Kernel-SVC based on TS-LDDMM representation is sensibly due to the misspecification of its regularization parameter.}

    \item \textbf{Classification benchmark on regularly sampled datasets, \Cref{appendix: settings_classification}:} We compare performances of a kernel support vector machine (SVC) algorithm based on TS-LDDMM representation with 3 state-of-the-art classification methods from shape analysis on 15 shape-based datasets (7 univariates \& 8 multivariates). \textbf{Results} in \Cref{table:wewin} show that the TS-LDDMM-based method outperforms other methods (best performances over 13 datasets), making TS-LDDMM representation relevant for time series shape analysis.
\end{enumerate}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/ts_ddmm_shooting.pdf}
    \caption{TS-LDDMM}
    \label{fig:ts-lddmm shooting}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/lddmm_shooting.pdf}
    \caption{LDDMM}
    \label{fig:lddmm shooting}
  \end{subfigure}
  \caption{Analysis of the two principal components (PC) related to mice' respiratory cycles before exposure for TS-LDDMM \Cref{fig:ts-lddmm shooting}, and LDDMM \Cref{fig:lddmm shooting}.
  In both cases, I) displays PC densities according to mice genotype and II) displays  deformations of the reference graph $S_0$ along each PC.}
  \label{fig:exp1}
\end{figure*}

\vspace{-1ex}
\subsection{Interpretability: analysis of respiratory behavior in mice}
\label{section:interpretability}
\vspace{-1ex}
%We consider a dataset composed of mouse respiratory time series before and after a drug injection.
% A complete presentation of this dataset is given in \Cref{appendix:mouse_dataset}.
%  The mouse are divided in two groups depending on their genotypes : \textit{colq} and \textit{wt}.
%  We aim to study the difference of respiratory shapes according to the genotype.
%   That is why, TS-LDDMM \eqref{eq:general_optimization_problem} is applied to derive the features representations $(\mathbf{\alpha}_j)_{j\in[N]}\in (\Rset^2)^N$ related to $N$=\sam{fill} respiratory time series coming from 14 mouse (7 \textit{colq} and \textit{wt}) before drug injection.
%   Then, a Principal Components Analysis (PCA) is performed on $(\mathbf{\alpha}_j)_{j\in[N]}$

@@

This experiment highlights the \textit{interpretability} of TS-LDDMM representation for studying the inter-individual variability in time series biomedical datasets. We consider a multivariate time series dataset that accounts for mice's nasal and thoracic airflow evolution when exposed to an irritant molecule altering respiratory functions \cite{nervo2019respiratory}.
The dataset is divided into two groups, one composed of 7 control mice (\textbf{wt}) and the other of 7 mice (\textbf{colq}) deficient in an enzyme involved in the control of respiration. For each mouse, airflows were recorded for 15 to 20 minutes before exposure to the irritant molecule and then for 35 to 40 minutes. A complete 
description of the dataset is given in the \Cref{appendix:mouse_dataset}.

\paragraph{Protocol.} For both experiments and representations (TS-LDDMM and LDDMM), we derive the reference respiratory cycle's graph $S_0$ and the representations $(\alpha_j)_{j\in[N]}$ related to $N$ respiratory cycles extracted according to the procedure \cite{germain2023unsupervised} by solving  \eqref{eq:general_optimization_problem}.
 Then, we perform a kernel PCA on the initial velocity field $\left(v_0(\alpha_j,S_0)\right)_{j\in[N]}\in \msv^{N_1}$ defined in \eqref{eq:def_v0} to breathing behavior.
  The first experiment includes $N_1 = 700$ respiratory cycles collected before exposure. The second experiment includes $N_2 = 1400$ respiratory cycles with 25\% (resp. 75\%) before (resp. after) exposure.
   \Cref{appendix: mice_exp_setting} describes methods settings.
    Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales. 
    \begin{table}[t]
      \centering
      \resizebox{\columnwidth}{!}{%
      \begin{tabular}{llrrrr}
        \toprule
         & \textbf{Dataset} & \textbf{Shape-FPCA (2024)} & \textbf{TCLR (2024)} & \textbf{LDDMM (2008)} & \textbf{TS-LDDMM (ours)} \\
        \midrule
        \multirow[c]{7}{*}{Univariate} & ArrowHead & 0.18 & 0.75 & \underline{0.84} & \textbf{0.91} \\
         & BME & 0.16 & \underline{1.00} & 0.82 & \textbf{1.00} \\
         & ECG200 & 0.40 & 0.67 & \textbf{0.81} & \underline{0.79} \\
         & FacesUCR & 0.08 & \underline{0.73} & 0.69 & \textbf{0.86} \\
         & GunPoint & 0.93 & \underline{0.97} & 0.83 & \textbf{1.00} \\
         & PhalangesOutlinesCorrect & 0.39 & \textbf{0.63} & \underline{0.53} & 0.52 \\
         & Trace & 0.55 & \underline{1.00} & 0.46 & \textbf{1.00} \\
        \cline{1-6}
        \multirow[c]{8}{*}{Multivariate} & ArticularyWordRecognition & -- & -- & \underline{0.98} & \textbf{1.00} \\
         & Cricket & -- & -- & \underline{0.77} & \textbf{0.93} \\
         & ERing & -- & -- & \underline{0.95} & \textbf{0.98} \\
         & Handwriting & -- & -- & \underline{0.22} & \textbf{0.44} \\
         & Libras & -- & -- & \underline{0.56} & \textbf{0.60} \\
         & NATOPS & -- & -- & \underline{0.82} & \textbf{0.82} \\
         & RacketSports & -- & -- & \textbf{0.83} & \underline{0.79} \\
         & UWaveGestureLibrary & -- & -- & \underline{0.72} & \textbf{0.81} \\
        \bottomrule
        \end{tabular}
        %
      }
      \label{table:wewin}
      \caption{All the state-of-the-art methods \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} mentionned in related works are compared to TS-LDDMM in \Cref{appendix: settings_classification}.}

    \end{table}
    \vspace{-1ex}
    \vspace{-1ex}\begin{wrapfigure}{r}{0.5\textwidth}
      \centering
      \begin{subfigure}[b]{0.15\textwidth}
        \centering
        \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
        \caption{colq cycle}
        \label{fig:colq-cycle}
      \end{subfigure}
      \begin{subfigure}[b]{0.15\textwidth}
        \centering
        \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
        \caption{PC1 vs PC2}
        \label{fig: pcs-scatter}
      \end{subfigure}
      \begin{subfigure}[b]{0.15\textwidth}
        \centering
        \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
        \caption{wt cycle}
        \label{fig:wt-reference}
      \end{subfigure}
      \caption{\Cref{fig:colq-cycle} is a \textbf{colq} respiratory cycle. \Cref*{fig: pcs-scatter} displays individual mouse reference respiratory cycle in the TS-LDDMM PC1-PC2 coordinates. \Cref{fig:wt-reference} is a referent respiratory cycle of a \textbf{wt} mouse learned with TS-LDDMM. }
      \vspace{-1em}
    \end{wrapfigure}
\paragraph{Breathing behaviors before exposure.}
 We focus on the analysis of the two first Principal Components (PC) for TS-LDDMM (\Cref{fig:ts-lddmm shooting}), and LDDMM (\Cref{fig:lddmm shooting}).
  A respiratory cycle is an inspiration (positive flow) followed by an expiration (negative flow); \Cref{fig:wt-reference} displays an example of \textbf{wt} mouse respiratory cycle.
   \Cref{fig:exp1} shows that principal components learned with TS-LDMM lead to deformations that remain respiratory cycles, while deformations learned with LDDMM are challenging to interpret as respiratory cycles.
   The LDDMM velocity field kernel is a Gaussian anisotropic kernel that accounts for time and space scales; however, the entanglement of time and space dimensions in the kernel does not guarantee the graph structure, and it makes the convergence of the method difficult (relative varifold loss error: TS-LDDMM: 0.06, LDDMM: 0.11).
    Regarding TS-LDDMM, \Cref{fig:ts-lddmm shooting} shows its principal components refer to different types of deformations.
     By interpreting \Cref{fig:ts-lddmm shooting}: PC1 accounts for time warping, PC2 expresses the trade-off between inspiration and expiration duration.
      Compared to \textbf{wt} mice, the distribution of \textbf{colq} mice feature along the PC1 axis has a heavy left tail, and the associated deformation (-2 $\sigma_{\text{PC}}$) shows an inspiration with two peaks.
       \Cref{fig:colq-cycle} shows an example of such respiratory cycles, which may be caused by motor impairment due to enzyme deficiency \cite{germain2023unsupervised}.
        Finally, \Cref{fig: pcs-scatter} shows that PC1 and PC2 capture the main differences between the two groups as their respective reference graphs $S^j$ are located in different parts of the space.

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.95\linewidth]{"./pictures/exp2.pdf"}
  \caption{Analysis of the first Principal Component (PC1) related to mice' respiratory cycles before and after exposure for TS-LDDMM. Figure A) displays PC densities according to mice genotype, Figure B) displays  deformations of the reference graph $S_0$ along PC1, and Figure C) displays respiratory cycles with respect to time in PC1 and PC3 coordinates}
  \label{fig:exp_2_PCA}
  \vspace{-1.5em}
\end{figure*}
\vspace{-1ex}
\paragraph{Breathing behaviors' evolution after exposure to the irritant molecule} 
TS-LDDMM representation features are learned on a dataset that includes respiratory cycles before and after exposure.
 \Cref{fig:exp_2_PCA} displays the first principal component PC since it encodes the effect of the irritant molecule as depicted in \Cref{fig:exp_2_PCA}-C) (the exposure occurs at 20 minutes). \Cref{fig:exp_2_PCA}-B) shows that the deformation (-1.5 $\sigma_{\text{PC}}$) leads to longer respiratory cycles that include a pause between inspiration and expiration as observed in \cite{germain2023unsupervised}.
  Conjointly, \Cref{fig:exp_2_PCA}-A) shows a bimodal distribution for \textbf{wt} mice whereas \textbf{wt}' distributions before exposure were unimodal (\Cref{fig:ts-lddmm shooting}). Indeed, the irritant molecule inhibits the action of the deficient enzyme, \textbf{wt} mice strongly react to the irritant molecule, whereas \textbf{colq} mice better adapt due to their deficiency \cite{germain2023unsupervised}.
%   \vspace{-1ex}
% \paragraph{Result summary.} By comparing the shape of individual respiratory cycles, we show that TS-LDDMM features can encode genotype distinctive breathing behaviors and their evolution after exposure to the irritant molecule contrary to LDDMM \cite{glaunes2008large}. 
% \vspace{-1ex}
%Secondly, we compare breathing behaviors before and after exposure to observe the impact of the irritant molecule.
% We follow the same procedure as for before exposure, but we take $N_2=1400$ respiratory cycles extracted according 
% to the procedure CITE....
% In \Cref{fig:exp_2_PCA}, we focus on the first Principal Component (PC) since it encodes the effect of the irritant molecule as demonstrated on Figure \ref{fig:exp_2_PCA}-C) (exposure at time 20).
% We observe on \Cref{fig:exp_2_PCA}-B) that after exposure the mouse have longuer breath such that the expiration is longuer than inspiration.
%  Moreover, we deduce from Figure \ref{fig:exp_2_PCA}-A) that \textbf{colq} are more constant in their breath compared to \textbf{wt} after exposure.
%
% In Figure REF, we 
% also display the reference respiratory cycle S0 and its deformations in the principal component directions.
%  Additionally, we learn each mouse's reference respiratory cycle and represent them in the first and third PC coordinates system in Figure REF.
